TCP Congestion Algorithms In Datacenters \\ \vspace{0.3cm} {\large LS Project}

### 2017-12-20 paper/paper.tex
spell check british english

Some congestion control algorithms are more aggressive than others, and it has
been reported that some algorithms cause others to back off excessively.
Tenants may, perhaps inadvertedly, cause inequal bandwidth distributions during
times of congestion.

@@

This report includes a performance and behavior analysis of different TCP congestion control algorithms in a (simulated) multitenant datacenter network. Based on the results, we will recommend datacenters on measures to take to keep providing a reliable environment for customers when multiple congestion control algoritms are active in the network, for example using Quality of Service or discouraging certain congestion control algorithms. The main research question is: {\it How can a datacenter provide a reliable and performant environment for customers when multiple TCP congestion control algorithms are used in the
network?}

@@

In 2010, the algorithm DCTCP was described by Mohammad Alizadeh et
al.\cite{dctcp-congestion-original} and published as
RFC8257\cite{dctcp-congestion} in 2017. DCTCP is optimized for datacenters and
provides a high-burst tolerance, low latency, and high throughput when the
datacenter has a small part of the buffer available\cite{dctcp-congestion}.

@@

In order to select the most common TCP congestion algorithms desk research will be performed. This desk research will show which algorithms are used more often than other algoritms by looking for the default algorithms in Operating Systems. Based on those results a list of TCP congestion algorithms will be composed.

@@

We will analyse one common scenario: multiple servers of different owners
inside a multitenant datacenter, sends data to clients outside the datacenter.
This could be users downloading software, the bursts while streaming a video
or movie, or large resources on a webpage.

@@

The tests conducted with the algorithms have two environment variables: packet loss and delay. Those environment variables are simulated using \texttt{netem}, which is a network emulation functionality in several Linux distributions \cite{linux-netem}. \texttt{netem} is applied to the incoming side of the interfaces of the clients to emulate the same packet loss and delay. Windows, which is tested as a server, does not support \texttt{netem}. Therefore the packet loss and delay are applied to the incoming interfaces of the clients. Packet loss and delay are chosen as the behavior of most TCP algorithms depends on those variables \cite{bbr-congestion}\cite{bic-tcp-congestion}\cite{compound-tcp-congestion}\cite{cubic-tcp-congestion}\cite{dctcp-congestion}. The values for the environment variables are chosen based on measurements performed by Verizon \cite{verizon-latency}.

@@

The dataset of Verizon \cite{verizon-latency} is used to calculate the packet loss and delay used in this research. The dataset includes global information and shows averages per month over the year 2017. The packet loss is calculated by taking the average of the Verizon Business Packet Delivery Statistics for Country Specific Metrics for each country. Those averages were inverted by subtracting the packet delivery percentage from 100, to show the actual packet loss in percentages and not the packet delivery. Based on the ascending sorted values a bar chart was created. This bar chart is shown in figure \ref{fig:packet-loss-chart}

@@

The tests will be performed using scripts written in Python\cite{git-lsproj}.
All nodes run a client, which performs actions as instructed by the controller.
The controller is responsible for controlling the tests: synchronizing time
among nodes, setting the TCP congestion algorithm of a node, and requesting one
node to open a listening port and instructing another to start sending at a
given time. Nodes report their measurements back to the controller.

@@

Desk research will be conducted to gather information related to traffic management. The goal is to find a mechanism that is capable of allocating bandwidth equally between customers of a datacenter. This is done by analyzing the results from the algorithm comparison and the initial desk research on traffic management.

@@

The BBR algorithm depends on parameters $BtlBw$ and $RT_{prop}$ \cite{bbr-congestion}. Those parameters are estimates defined by the BBR algorithm. $BtlBw$ specifies the bandwidth at the slowest link in each direction and $RT_{prop}$ is the round-trip propagation time. $RT_{prop} = RTT - queuing delay - processing delay$, which results in the minimum amount of time for round trip propagation in case there are no queueing or processing delays. Using those two parameters, the Bandwidth Delay Product (BDP) is calculated, which is the maximum possible amount of data being sent in a network: $BDP = BtlBw * RT_{prop}$. As long as the amount of data in transit is less than the $BDP$ and $BtlBw$, the delivery rate is increased. If the amount of data in transit is equal to $BtlBw$ the delivery rate can not go up anymore. The $RTT$ can never be lower than $RT_{prop}$... explain more on delay perhaps?.

The CTCP algorithm is a combination of a loss-based and delay-based congestion protocol \cite{compound-tcp-congestion}. For the loss-based component, the variable conventional congestion window, $cwnd$, was introduced. For the delay-based component, the variable delay window, $dwnd$, was introduced. The TCP sending window is determined from: $win = min(cwnd + dwnd, awnd)$, where $awnd$ is the window advertised by the receiver. For $cwnd$ on the arrival of an ACK: $cwnd = cwnd + 1/win$. The $dwnd$ variable is initially set to zero and is only set when the congestion avoidance is active. $dwnd$ is calculated differently for several scenarios. If the network path is underutilized: $dwnd(t+1) = dwnd(t) + \alpha*dwnd(t)^{k} - 1$. When the network path is congested: $dwnd(t+1) = dwnd(t) - \eta*diff$. When packet loss occurs: $dwnd(t+1) = dwnd(t)(1-\beta) - cwnd/2$. $k$, $\alpha$, $\eta$ and $\beta$ are tunable parameters for optimization of the algorithm. $\eta$ defines the rapidness of reduction of the window when congestion is detected. $\beta$ is the factor used for the multiplicative decrease after loss is detected. $diff$ is calculated from the expected throughput and the actual throughput, which are again calculated from the window size and Round Trip Time (RTT).

@@

\begin{enumerate}
	\item Simple Marking at the Switch. Based on the marking treshold $K$ a packet is marked with the CE flag. This only occurs when the queue occupancy is greater than K at arrival time. The marking treshold $K$ specifies the minimum value of queue occupancy.
	\item ECN-Echo at receiver. The receiver running the DCTCP algorithm ACKs every packet and sets the ECN-Echo flag only if the packet was marked with CE. The ECN-Echo flag is used to notify the sender of congestion in the network.
	\item Controller at the sender. The sender has an estimate of the fraction of packets that are marked with CE. This is based on the actual fraction of packets that were marked, the weight given to the new estimate compared to the past weight. This results in $\alpha$. $\alpha$ is used to indicate the congestion on a network with a maximum of value 1, indicating a congested network, and a minimum value of 0, indicating no congestion in the network. DCTCP cuts the window size using $\alpha$: $cwnd = cwnd * (1 - \alpha/2)$. This results in a low queue length, but ensuring high throughput.\\
\end{enumerate}

@@

The CUBIC algorithm is a successor of the BIC algorithm \cite{cubic-tcp-congestion}. In case of a loss event $W_{max}$, which is the maximum window size, is registered. This is followed by a multiplicative decrease of the congestion window by factor $\beta$. If an ACK in congestion avoidance is received and the current window size $cwnd$ is less than $W_{max}$, the algorithm uses the concave profile to increase the window size according to the formula: $cwnd = \frac{W(t+RTT)-cwnd}{cwnd}$. If $cwnd$ is larger than $W_{max}$, the convex profile is used to increase the window size according to the same formula as the concave profile. In both formulas $t$ is the time elapsed from the last window reducation.

@@

With the use of a hypothetical TCP congestion control algorithm that is unfair towards other flows, the bandwidth allocation to one customer would be higher than to other customers. When a TCP flow shows an aggressive increase and keeps pushing data over the network, other (fair) algorithms will reduce the transmission rate as packet loss occurs. One important condition for this unfair congestion control algorithm is that it should not be agressive to the extent of causing a denial of service.

To mitigate this, the bandwidth usage per server could be monitored, prioritizing the traffic sent by some servers. The solution uses traffic shaping to control the priority of the traffic per server:

\begin{enumerate}
	\item The monitoring solution measures periodically whether the bandwidth in the link is fully utilized. If this is the case, the next step will be performed. If not, it continues monitoring the link utilization.
	\item The monitoring solution calculates periodically per server whether its bandwidth usage deviates more than $\alpha$ from the average bandwidth usage. $\alpha$ is deviated bandwidth and is set according to an acceptable deviation value. The following formula is used to calculate the standard deviation: $s_{i} = \sqrt{\frac{\sum_{i=1}^N (x_i - \overline{x})^2}{N-1} }$. $N$ is the amount of active servers on the link. $\overline{x}$ is the average amount of bandwidth for all active servers on the link. $x_i$ is the bandwidth of the current server and $s_{i}$ shows the standard deviation in bandwidth for server $i$. If $s_{i}$ is a negative value, no priority change will be applied as the bandwidth usage is lower than average for the server. If $s_{i} <= \alpha$, then no priority change will be applied. If $s_{i} > \alpha$, the priority will be changed to a lower value.
	\item The monitoring solution provides information of the servers where the priority will be set to a lower value to the traffic shaping controller.
	\item The traffic shaping controller sets the priority for the servers to a lower value, so that the traffic coming from the servers that use $\alpha$ more than average bandwidth has less preference than traffic closer to the average bandwidth and lower.
	\item The monitoring solution periodically checks if the bandwidth in the link is still fully utilized. If this is not the case, the traffic shaping policy is reversed again by setting the priority to the default value. If this is the case, the periodical calculations of server bandwidth usage are continued.
\end{enumerate}

@@

Without any concurrent flows, Windows does reach gigabit speed. On different
hardware, it showed the same, recessive behaviour. We think this might be
caused by the way Python socket operations are handled in Windows. When
repeating the test with \texttt{iperf}, which is written in C, this behaviour
no longer occurs. Even when using the Python equivalent of
\texttt{while(true)\{ socket.send(data); \}}, the bandwidth allocation did not
change. We theorize that the difference is caused by something in the
\texttt{send} call. An alternative theory is that Microsoft Windows for
workstations have a different implementation than Microsoft Windows Server
systems.

@@

Our main research question is: {\it How can a datacenter provide a reliable and performant environment for customers when multiple TCP congestion control algorithms are used in the network?}

>>>>

Some congestion control algorithms are more aggressive than others, and it has
been reported that some algorithms cause others to back off excessively.
Tenants may, perhaps inadvertently, cause unequal bandwidth distributions during
times of congestion.

@@

This report includes a performance and behaviour analysis of different TCP congestion control algorithms in a (simulated) multitenant datacenter network. Based on the results, we will recommend datacenters on measures to take to keep providing a reliable environment for customers when multiple congestion control algorithms are active in the network, for example using Quality of Service or discouraging certain congestion control algorithms. The main research question is: {\it How can a datacenter provide a reliable and performing environment for customers when multiple TCP congestion control algorithms are used in the
network?}

@@

In 2010, the algorithm DCTCP was described by Mohammad Alizadeh et
al.\cite{dctcp-congestion-original} and published as
RFC8257\cite{dctcp-congestion} in 2017. DCTCP is optimised for datacenters and
provides a high-burst tolerance, low latency, and high throughput when the
datacenter has a small part of the buffer available\cite{dctcp-congestion}.

@@

In order to select the most common TCP congestion algorithms desk research will be performed. This desk research will show which algorithms are used more often than other algorithms by looking for the default algorithms in Operating Systems. Based on those results a list of TCP congestion algorithms will be composed.

@@

We will analyse one common scenario: multiple servers of different owners
inside a multitenant datacenter, sends data to clients outside the datacenter.
This could be users downloading software, the bursts while streaming a video
or movie, or large resources on a web page.

@@

The tests conducted with the algorithms have two environment variables: packet loss and delay. Those environment variables are simulated using \texttt{netem}, which is a network emulation functionality in several Linux distributions \cite{linux-netem}. \texttt{netem} is applied to the incoming side of the interfaces of the clients to emulate the same packet loss and delay. Windows, which is tested as a server, does not support \texttt{netem}. Therefore the packet loss and delay are applied to the incoming interfaces of the clients. Packet loss and delay are chosen as the behaviour of most TCP algorithms depends on those variables \cite{bbr-congestion}\cite{bic-tcp-congestion}\cite{compound-tcp-congestion}\cite{cubic-tcp-congestion}\cite{dctcp-congestion}. The values for the environment variables are chosen based on measurements performed by Verizon \cite{verizon-latency}.

@@

The data set of Verizon \cite{verizon-latency} is used to calculate the packet loss and delay used in this research. The data set includes global information and shows averages per month over the year 2017. The packet loss is calculated by taking the average of the Verizon Business Packet Delivery Statistics for Country Specific Metrics for each country. Those averages were inverted by subtracting the packet delivery percentage from 100, to show the actual packet loss in percentages and not the packet delivery. Based on the ascending sorted values a bar chart was created. This bar chart is shown in figure \ref{fig:packet-loss-chart}

@@

The tests will be performed using scripts written in Python\cite{git-lsproj}.
All nodes run a client, which performs actions as instructed by the controller.
The controller is responsible for controlling the tests: synchronising time
among nodes, setting the TCP congestion algorithm of a node, and requesting one
node to open a listening port and instructing another to start sending at a
given time. Nodes report their measurements back to the controller.

@@

Desk research will be conducted to gather information related to traffic management. The goal is to find a mechanism that is capable of allocating bandwidth equally between customers of a datacenter. This is done by analysing the results from the algorithm comparison and the initial desk research on traffic management.

@@

The BBR algorithm depends on parameters $BtlBw$ and $RT_{prop}$ \cite{bbr-congestion}. Those parameters are estimates defined by the BBR algorithm. $BtlBw$ specifies the bandwidth at the slowest link in each direction and $RT_{prop}$ is the round-trip propagation time. $RT_{prop} = RTT - queuing delay - processing delay$, which results in the minimum amount of time for round trip propagation in case there are no queuing or processing delays. Using those two parameters, the Bandwidth Delay Product (BDP) is calculated, which is the maximum possible amount of data being sent in a network: $BDP = BtlBw * RT_{prop}$. As long as the amount of data in transit is less than the $BDP$ and $BtlBw$, the delivery rate is increased. If the amount of data in transit is equal to $BtlBw$ the delivery rate can not go up anymore. The $RTT$ can never be lower than $RT_{prop}$... explain more on delay perhaps?.

The CTCP algorithm is a combination of a loss-based and delay-based congestion protocol \cite{compound-tcp-congestion}. For the loss-based component, the variable conventional congestion window, $cwnd$, was introduced. For the delay-based component, the variable delay window, $dwnd$, was introduced. The TCP sending window is determined from: $win = min(cwnd + dwnd, awnd)$, where $awnd$ is the window advertised by the receiver. For $cwnd$ on the arrival of an ACK: $cwnd = cwnd + 1/win$. The $dwnd$ variable is initially set to zero and is only set when the congestion avoidance is active. $dwnd$ is calculated differently for several scenarios. If the network path is underutilised: $dwnd(t+1) = dwnd(t) + \alpha*dwnd(t)^{k} - 1$. When the network path is congested: $dwnd(t+1) = dwnd(t) - \eta*diff$. When packet loss occurs: $dwnd(t+1) = dwnd(t)(1-\beta) - cwnd/2$. $k$, $\alpha$, $\eta$ and $\beta$ are tunable parameters for optimisation of the algorithm. $\eta$ defines the rapidness of reduction of the window when congestion is detected. $\beta$ is the factor used for the multiplicative decrease after loss is detected. $diff$ is calculated from the expected throughput and the actual throughput, which are again calculated from the window size and Round Trip Time (RTT).

@@

\begin{enumerate}
	\item Simple Marking at the Switch. Based on the marking threshold $K$ a packet is marked with the CE flag. This only occurs when the queue occupancy is greater than K at arrival time. The marking threshold $K$ specifies the minimum value of queue occupancy.
	\item ECN-Echo at receiver. The receiver running the DCTCP algorithm ACKs every packet and sets the ECN-Echo flag only if the packet was marked with CE. The ECN-Echo flag is used to notify the sender of congestion in the network.
	\item Controller at the sender. The sender has an estimate of the fraction of packets that are marked with CE. This is based on the actual fraction of packets that were marked, the weight given to the new estimate compared to the past weight. This results in $\alpha$. $\alpha$ is used to indicate the congestion on a network with a maximum of value 1, indicating a congested network, and a minimum value of 0, indicating no congestion in the network. DCTCP cuts the window size using $\alpha$: $cwnd = cwnd * (1 - \alpha/2)$. This results in a low queue length, but ensuring high throughput.\\
\end{enumerate}

@@

The CUBIC algorithm is a successor of the BIC algorithm \cite{cubic-tcp-congestion}. In case of a loss event $W_{max}$, which is the maximum window size, is registered. This is followed by a multiplicative decrease of the congestion window by factor $\beta$. If an ACK in congestion avoidance is received and the current window size $cwnd$ is less than $W_{max}$, the algorithm uses the concave profile to increase the window size according to the formula: $cwnd = \frac{W(t+RTT)-cwnd}{cwnd}$. If $cwnd$ is larger than $W_{max}$, the convex profile is used to increase the window size according to the same formula as the concave profile. In both formulas $t$ is the time elapsed from the last window reduction.

@@

With the use of a hypothetical TCP congestion control algorithm that is unfair towards other flows, the bandwidth allocation to one customer would be higher than to other customers. When a TCP flow shows an aggressive increase and keeps pushing data over the network, other (fair) algorithms will reduce the transmission rate as packet loss occurs. One important condition for this unfair congestion control algorithm is that it should not be aggressive to the extent of causing a denial of service.

To mitigate this, the bandwidth usage per server could be monitored, prioritising the traffic sent by some servers. The solution uses traffic shaping to control the priority of the traffic per server:

\begin{enumerate}
	\item The monitoring solution measures periodically whether the bandwidth in the link is fully utilised. If this is the case, the next step will be performed. If not, it continues monitoring the link utilisation.
	\item The monitoring solution calculates periodically per server whether its bandwidth usage deviates more than $\alpha$ from the average bandwidth usage. $\alpha$ is deviated bandwidth and is set according to an acceptable deviation value. The following formula is used to calculate the standard deviation: $s_{i} = \sqrt{\frac{\sum_{i=1}^N (x_i - \overline{x})^2}{N-1} }$. $N$ is the amount of active servers on the link. $\overline{x}$ is the average amount of bandwidth for all active servers on the link. $x_i$ is the bandwidth of the current server and $s_{i}$ shows the standard deviation in bandwidth for server $i$. If $s_{i}$ is a negative value, no priority change will be applied as the bandwidth usage is lower than average for the server. If $s_{i} <= \alpha$, then no priority change will be applied. If $s_{i} > \alpha$, the priority will be changed to a lower value.
	\item The monitoring solution provides information of the servers where the priority will be set to a lower value to the traffic shaping controller.
	\item The traffic shaping controller sets the priority for the servers to a lower value, so that the traffic coming from the servers that use $\alpha$ more than average bandwidth has less preference than traffic closer to the average bandwidth and lower.
	\item The monitoring solution periodically checks if the bandwidth in the link is still fully utilised. If this is not the case, the traffic shaping policy is reversed again by setting the priority to the default value. If this is the case, the periodical calculations of server bandwidth usage are continued.
\end{enumerate}

@@

Without any concurrent flows, Windows does reach gigabit speed. On different
hardware, it showed the same, recessive behaviour. We think this might be
caused by the way Python socket operations are handled in Windows. When
repeating the test with \texttt{iperf}, which is written in C, this behaviour
no longer occurs. Even when using the Python equivalent of
\texttt{while(true)\{ socket.send(data); \}}, the bandwidth allocation did not
change. We theorise that the difference is caused by something in the
\texttt{send} call. An alternative theory is that Microsoft Windows for
workstations have a different implementation than Microsoft Windows Server
systems.

@@

Our main research question is: {\it How can a datacenter provide a reliable and performing environment for customers when multiple TCP congestion control algorithms are used in the network?}
